---
title: "IVEA tutorial: predicting enhancer–gene interactions in K562 (chr22)"
author: "IVEA"
output:
  rmarkdown::html_document:
    toc: true
    toc_depth: 2
    df_print: kable
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  fig.align = "center",
  message = FALSE,
  warning = FALSE
)
options(width = 100)
```

This tutorial runs the full IVEA pipeline on the bundled **K562 chr22** example, inspects the
predicted promoter/enhancer activities and enhancer–gene interaction scores, and then walks
through one gene — **TPST2** — to read its predicted regulatory landscape and the *uncertainty*
IVEA reports for each estimate.

IVEA estimates promoter and enhancer *activities* by variational Bayesian inference and scores
enhancer–gene regulatory interactions. Method and evaluation: Kimura et al., *Bioinformatics
Advances* 2024;4(1):vbae118, <https://doi.org/10.1093/bioadv/vbae118>.

```{r libraries}
library(IVEA)
library(ggplot2)
theme_set(theme_minimal(base_size = 12))
```

The example inputs live outside the installed package, in the source tree (`example/output/`
and `reference/`). Two working directories occur when this article is rendered: standalone
`rmarkdown::render()` runs from this file's folder (`vignettes/articles/`), while pkgdown runs
from the package root. We point `repo` at the root in either case, then build the paths from it.

```{r locate-data}
repo   <- if (file.exists("DESCRIPTION")) "." else "../.."
outdir <- file.path(repo, "example", "output")
stopifnot(dir.exists(outdir))
```

# 1. Run IVEA on chr22

IVEA's public API is three steps: `load_data()` reads and filters the inputs into the `ls_x`
data list, `run_variational_bayes()` runs the coordinate-ascent inference, and `make_summary()`
turns the final estimates into tidy output tables. This is exactly what the command-line driver
`scripts/run_IVEA.R` does; here we call the functions directly.

We assemble one configuration list whose fields mirror the driver's options (and the settings in
`example/commands_example.sh`). The scaling hyperparameters `alpha_pro = 80` and `alpha_enh = 10`
match the example.

```{r config}
cfg <- list(
  # Enhancer activity prior model:
  #   TRUE  -> IVEA_nolE (relative enhancer length dropped from the prior). This is the
  #            driver default and the better-performing model reported in the paper.
  #   FALSE -> original IVEA (uses relative enhancer length).
  no_le_enh = TRUE,

  # Gamma-prior hyperparameters
  alpha_pro = 80, alpha_enh = 10,
  alpha_o = 0.01, beta_o = 1e-4,
  alpha_k = 1e-3, beta_k = 1e-3,

  # Iteration control
  max_iteration = 500, stop_criterion = 0.005,

  # Chromatin contact
  contacts = file.path(outdir, "enhancer-gene_contacts.chr22.txt"),
  hic_contact_colname = NULL,   # use the last (power-law) contact column
  window = 5e6,

  # Gene expression (RNA-seq)
  rnas = file.path(outdir, "gene_expressions.txt"),
  rna_id_colname = "gene_name",
  rna_length_colname = "effective_length",
  rna_count_colname = "expected_count",
  rna_cutoff_colname = "TPM",
  rna_cutoff_value = 8,

  # Regulatory elements + burst sizes
  enhancers  = file.path(outdir, "enhancer_elements.txt"),
  promoters  = file.path(outdir, "gene_promoter_elements.txt"),
  burst_sizes = file.path(repo, "reference", "gene_burst_sizes.hg19.txt")
)
```

**Two models, one switch.** IVEA ships two variants of the enhancer-activity prior, selected by
`no_le_enh`. `TRUE` gives `IVEA_nolE` (the default, and the model the paper found predicts best);
`FALSE` gives the original `IVEA`, which also uses each enhancer's relative length. We use the
recommended `TRUE` here. (The pre-computed files in `example/output/` were generated with
`FALSE`, so the numbers below will differ slightly from those files — that is expected.)

```{r run, results = "hide"}
ls_x <- load_data(cfg, "chr22", NULL)
ls_z <- run_variational_bayes(cfg, ls_x)
ls_out <- make_summary(ls_x, ls_z)
```

The inference analysed **`r length(ls_x$gene_name)` genes** and
**`r length(ls_x$enh_name)` enhancers** on chr22, converging after
**`r length(ls_z$k) - 1` iterations** (relative change in enhancer activities < `r cfg$stop_criterion`).
`make_summary()` returns four tables:

- `ls_out$info_pair` — one row per enhancer–gene pair, with the interaction `score` and its
  components (score-sorted);
- `ls_out$bedpe_pair` — the same pairs in BEDPE format;
- `ls_out$bed_pro` / `ls_out$bed_enh` — per-promoter / per-enhancer activity **with uncertainty**
  (posterior mean, 2.5/50/97.5 percentiles, SD, and Gamma `shape`/`rate`).

# 2. Inspect enhancer activity

`ls_out$bed_enh` holds each enhancer's estimated activity and the uncertainty of that estimate.

```{r enh-head}
head(ls_out$bed_enh[, c("name", "activity", "activity_025", "activity_975", "sd_activity", "shape", "rate")])
```

Activities span two–three orders of magnitude, so we view them on a log scale.

```{r enh-hist}
ggplot(ls_out$bed_enh, aes(activity)) +
  geom_histogram(bins = 40, fill = "#4C72B0", colour = "white", linewidth = 0.2) +
  scale_x_log10() +
  labs(x = "Enhancer activity (log scale)", y = "Number of enhancers",
       title = "Distribution of estimated enhancer activities (chr22)")
```

Each activity is the mean of a posterior **Gamma** distribution; the `activity_025`/`activity_975`
columns are its 2.5/97.5 percentiles and `sd_activity` its standard deviation. We return to what
drives that uncertainty in section 6.

# 3. Inspect enhancer–gene interactions

`ls_out$info_pair` is sorted by descending `score`. The score combines enhancer activity, chromatin
contact, and promoter activity; `strength` and `contribution` are intermediate quantities and
`distance` is the enhancer-to-TSS distance in bp.

```{r pair-head}
knitr::kable(head(ls_out$info_pair[, c("gene", "enhancer", "distance", "contact", "strength", "contribution", "score")]), row.names = FALSE)
```

Interaction scores decay with genomic distance — nearby enhancers dominate, with a long tail of
weaker distal links:

```{r score-vs-distance}
ggplot(ls_out$info_pair, aes(distance + 1, score)) +
  geom_point(alpha = 0.5, size = 1.0, colour = "#4C72B0") +
  scale_x_log10(labels = function(x) paste0(x / 1e3, " kb")) +
  labs(x = "Enhancer–TSS distance (log scale)", y = "Interaction score",
       title = "Interaction score decays with distance")
```

# 4. A worked example: TPST2

We focus on **TPST2**. Its promoter activity (with a 95% credible interval) comes from
`ls_out$bed_pro`:

```{r tpst2-promoter}
gene <- "TPST2"
df_pro <- ls_out$bed_pro[ls_out$bed_pro$name == gene, ]
knitr::kable(df_pro[, c("name", "p_end", "activity", "activity_025", "activity_975", "sd_activity")], row.names = FALSE)
```

Its enhancer–gene pairs, ranked by score:

```{r tpst2-pairs}
df_pair <- ls_out$info_pair[ls_out$info_pair$gene == gene, ]
df_pair <- df_pair[order(-df_pair$score), ]
knitr::kable(head(df_pair[, c("enhancer", "distance", "contact", "strength", "score")], 10), row.names = FALSE)
```

```{r tpst2-coords, echo = FALSE}
# Parse "chr:start-end" enhancer strings into numeric midpoints for plotting.
m <- regmatches(df_pair$enhancer, regexec("^(chr[0-9XYM]+):([0-9]+)-([0-9]+)$", df_pair$enhancer))
df_pair$e_start <- as.numeric(vapply(m, `[`, "", 3))
df_pair$e_end   <- as.numeric(vapply(m, `[`, "", 4))
df_pair$e_mid   <- (df_pair$e_start + df_pair$e_end) / 2
tss <- df_pair$gene_tss[1]
```

TPST2 has **`r nrow(df_pair)` candidate pairs**; a handful of proximal enhancers carry most of the
predicted regulatory input, tailing off to weak distal links.

# 5. The predicted regulatory landscape

Before drawing the landscape, we bring in **ground truth**. IVEA bundles a set of
CRISPRi-validated K562 enhancer–gene pairs under `evaluation_data/`
(`K562.combined.{positive,negative}.hg19.bedpe`), combining screens from Fulco et al. (2019),
Gasperini et al. (2019), and Schraivogel et al. (2020): regions where perturbing a candidate
enhancer did (*positive*) or did not (*negative*) change the target gene's expression. For TPST2
that panel contains one positive pair and five negatives (all from Gasperini et al. 2019). We read
them straight from those files, map each validated region to the IVEA enhancer it overlaps, and
read off that enhancer's rank among TPST2's `r nrow(df_pair)` predicted pairs.

```{r validated-pairs}
n_top <- 10   # "top 10" throughout this section

# Read the CRISPRi-validated pairs for this gene from the bundled evaluation data.
# Each file is a headerless 7-column BEDPE: enhancer chr/start/end, TSS chr/start/end, gene.
read_validated <- function(file, class) {
  df <- read.table(file.path(repo, "evaluation_data", file), sep = "\t",
                  col.names = c("chr", "g_start", "g_end", "g_chr", "tss1", "tss2", "gene"))
  df <- df[df$gene == gene, c("g_start", "g_end")]
  if (nrow(df) == 0) return(NULL)
  data.frame(class = class, df)
}
df_validated <- rbind(
  read_validated("K562.combined.positive.hg19.bedpe", "positive"),
  read_validated("K562.combined.negative.hg19.bedpe", "negative")
)
df_validated$g_mid <- (df_validated$g_start + df_validated$g_end) / 2

# Rank TPST2's pairs, then map each validated region to the overlapping IVEA enhancer.
df_pair$rank <- seq_len(nrow(df_pair))   # df_pair is already sorted by descending score
map_region <- function(s, e) {
  hit <- df_pair[df_pair$e_start <= e & df_pair$e_end >= s, ]
  if (nrow(hit) == 0) return(data.frame(enhancer = NA_character_, rank = NA_integer_, score = NA_real_))
  hit[which.min(hit$rank), c("enhancer", "rank", "score")]
}
df_validated <- cbind(df_validated, do.call(rbind, Map(map_region, df_validated$g_start, df_validated$g_end)))
df_validated$in_top10 <- !is.na(df_validated$rank) & df_validated$rank <= n_top

knitr::kable(df_validated[, c("class", "g_start", "g_end", "enhancer", "rank", "in_top10")], row.names = FALSE)
```

The single **positive** maps to a **rank-`r df_validated$rank[df_validated$class == "positive"]`**
enhancer — comfortably inside the top `r n_top`. Every **negative** falls outside: two map to
ranks `r paste(sort(df_validated$rank[df_validated$class == "negative"]), collapse = " and ")`, and the
other three overlap no called enhancer at all. IVEA's ranking therefore separates the validated
positive from the validated negatives.

We draw the locus as a track: the TSS is marked at the bottom, each enhancer is a lollipop at its
genomic midpoint with height equal to its interaction score, and an arc connects the TSS to each
enhancer with width scaled by score. Colour encodes score (a magnitude, so a single sequential
hue). The enhancer overlapping the CRISPRi-validated **positive** is ringed.

```{r landscape, fig.height = 5}
df_top <- df_pair[order(-df_pair$score), ][seq_len(min(n_top, nrow(df_pair))), ]
df_pos_hit <- df_top[df_top$enhancer %in% df_validated$enhancer[df_validated$class == "positive"], ]

ggplot(df_top, aes(x = e_mid / 1e6)) +
  # arcs TSS -> enhancer, weighted by score
  geom_curve(aes(x = tss / 1e6, y = 0, xend = e_mid / 1e6, yend = 0, linewidth = score),
             curvature = -0.2, colour = "#4C72B0", lineend = "round") +
  # lollipop stems + heads
  geom_segment(aes(xend = e_mid / 1e6, y = 0, yend = score), colour = "grey70", linewidth = 0.3) +
  geom_point(aes(y = score, colour = score), size = 2.3) +
  # ring + label the CRISPRi-validated positive
  geom_point(data = df_pos_hit, aes(y = score), shape = 21, size = 4.5, stroke = 1.1,
             colour = "#0072B2", fill = NA) +
  geom_text(data = df_pos_hit, aes(y = score, label = "CRISPRi +"),
            vjust = 0.5, hjust = -0.2, size = 3, fontface = "bold", colour = "#0072B2") +
  # TSS marker
  geom_point(aes(x = tss / 1e6, y = 0), shape = 17, size = 3, colour = "#C44E52") +
  annotate("text", x = tss / 1e6, y = 0, label = paste0("  ", gene, " TSS"),
           hjust = 0, vjust = 1.6, size = 3.4, colour = "#C44E52") +
  # direct-label the enhancer scores
  geom_text(data = df_top, aes(y = score, label = sprintf("%.3f", score)),
            vjust = -0.9, size = 3, colour = "grey20") +
  scale_colour_viridis_c(option = "D", name = "Score") +
  scale_linewidth(range = c(0.3, 1.5), guide = "none") +
  labs(x = "chr22 position (Mb)", y = "Interaction score",
       title = paste0("Top ", n_top, " predicted regulatory landscape of ", gene)) +
  theme(panel.grid.minor = element_blank())
```

The dominant enhancers sit closest to the TSS; the score falls off with distance, matching the
genome-wide pattern in section 3.

To see the positive and negatives together against the **top-10 cutoff**, we plot every candidate
enhancer within **1 Mb of the TSS**, coloured by interaction score exactly as in the landscape
above, and overlay the validated regions. The dashed line is the score of the 10th-ranked pair;
the CRISPRi-validated positive (ringed, as in the landscape) sits above it, while the negatives
(crossed) fall below (or at score 0 where no enhancer was called).

```{r validation-plot, fig.height = 4.5}
thr <- df_top$score[n_top]                              # top-10 score cutoff
win <- tss + c(-1e6, 1e6)                            # TPST2 TSS +/- 1 Mb
bg  <- df_pair[df_pair$e_mid >= win[1] & df_pair$e_mid <= win[2], ] # candidate enhancers in that window

df_pos_hit <- df_top[df_top$enhancer %in% df_validated$enhancer[df_validated$class == "positive"], ]
df_neg <- df_validated[df_validated$class == "negative", ]
df_neg$plot_score <- ifelse(is.na(df_neg$score), 0, df_neg$score)  # 0 where no enhancer was called

ggplot(bg, aes(e_mid / 1e6, score)) +
  # candidate enhancers, coloured by interaction score (as in the landscape)
  geom_point(aes(colour = score), size = 1.7) +
  geom_hline(yintercept = thr, linetype = "dashed", colour = "grey50") +
  annotate("text", x = win[1] / 1e6, y = thr, label = "top-10 cutoff",
           hjust = 0, vjust = -0.5, size = 3, colour = "grey45") +
  # CRISPRi-validated negatives
  geom_point(data = df_neg, aes(g_mid / 1e6, plot_score), shape = 4, size = 3, stroke = 1.2,
             colour = "#D55E00") +
  # CRISPRi-validated positive: same blue ring + label as the landscape figure
  geom_point(data = df_pos_hit, aes(e_mid / 1e6, score), shape = 21, size = 4.5, stroke = 1.1,
             colour = "#0072B2", fill = NA) +
  geom_text(data = df_pos_hit, aes(e_mid / 1e6, score, label = "CRISPRi +"),
            vjust = 0.5, hjust = -0.2, size = 3, fontface = "bold", colour = "#0072B2") +
  # TSS marker
  annotate("point", x = tss / 1e6, y = 0, shape = 17, size = 3, colour = "#C44E52") +
  annotate("text", x = tss / 1e6, y = 0, label = paste0("  ", gene, " TSS"),
           hjust = 0, vjust = 1.6, size = 3.2, colour = "#C44E52") +
  scale_colour_viridis_c(option = "D", name = "Score") +
  labs(x = "chr22 position (Mb)", y = "Interaction score",
       title = "CRISPRi-validated pairs vs predicted score",
       subtitle = "Positive (ring) clears the top-10 cutoff; negatives (crosses) do not") +
  theme(panel.grid.minor = element_blank())
```

# 6. The predicted enhancer activity

Finally we look at the **activity** estimates (not the interaction scores) of the enhancers that
make up TPST2's top-10 interactions, positioned along the chromosome. Each point is the posterior
**median** (`activity_050`); the vertical range is the **2.5–97.5 percentile** credible interval
(`activity_025`–`activity_975`). Colour carries the interaction score, so you can see how activity
and predicted regulatory role line up across the locus.

```{r activity-position, fig.height = 4.5}
df_top_enh <- merge(df_top[, c("enhancer", "rank", "score", "e_start", "e_end", "e_mid")],
                 ls_out$bed_enh[, c("name", "activity_025", "activity_050", "activity_975")],
                 by.x = "enhancer", by.y = "name")
df_pos_enh <- df_top_enh[df_top_enh$enhancer %in% df_validated$enhancer[df_validated$class == "positive"], ]

ggplot(df_top_enh, aes(x = e_mid / 1e6)) +
  geom_vline(xintercept = tss / 1e6, linetype = "dotted", colour = "#C44E52") +
  annotate("text", x = tss / 1e6, y = max(df_top_enh$activity_975), label = paste0(" ", gene, " TSS"),
           hjust = 0, size = 3, colour = "#C44E52") +
  geom_linerange(aes(ymin = activity_025, ymax = activity_975), colour = "grey65", linewidth = 1.5) +
  geom_point(aes(y = activity_050, colour = score), shape = 15, size = 3) +
  coord_cartesian(ylim = c(0, max(df_top_enh$activity_975))) +
  # ring the CRISPRi-validated positive
  geom_point(data = df_pos_enh, aes(y = activity_050), shape = 21, size = 5.5, stroke = 1.1,
             colour = "#0072B2", fill = NA) +
  scale_colour_viridis_c(option = "D", name = "Interaction\nscore") +
  labs(x = "chr22 position (Mb)",
       y = "Enhancer activity (2.5–97.5 %ile; point = median)",
       title = paste0("Enhancer activity across the top-", n_top, " ", gene, " interactions"),
       subtitle = "Ringed = the CRISPRi-validated positive enhancer") +
  theme(panel.grid.minor.x = element_blank())
```

The intervals show that these top enhancers are estimated with differing confidence. Note that
activity and interaction score need not track each other: the CRISPRi-validated positive has the
**highest activity** in the panel yet ranks `r df_validated$rank[df_validated$class == "positive"]`,
because the score weighs activity by chromatin contact — the proximal enhancers rank higher on more
modest activity thanks to their closeness to the TSS.

# Session information

```{r session, echo = FALSE}
sessionInfo()
```
